Simulating the Linux page cache in distributed applications with WRENCH

### 2020-12-09 paper_io_sim.tex
Henri's pass on abstract and Section 1

\title{Simulating the Linux page cache in distributed applications with WRENCH}

\author{\IEEEauthorblockN{Hoang-Dung Do$^1$, Val\'erie Hayot-Sasson$^1$, Rafael Ferreira da Silva$^3$, Henri Casanova$^2$, Tristan Glatard$^1$
  }\\
  \IEEEauthorblockA{
    $^1$Department of Computer Science and Software Engineering, Concordia University, Montreal, Canada\\
     $^2$Department of Information and Computer Sciences, University of Hawai‘i at Manoa, USA\\
     $^3$Department of Computer Science, University of Southern California, Los Angeles, USA
  }
}

@@

        The rapid emergence of Big Data seen in recent years has resulted in a 
        growing need for efficient data storing and optimized data analysis techniques. 
        Despite relying on large computing infrastructures with powerful computing capability 
        to process these large datasets, an I/O bottleneck remains. 
        One widely-used approach to minimize the costs of data transfers during
        regular use is through the employment of the Linux page cache.
        While many simulation tools have been developed to 
        support studies in computing systems optimization, page cache simulation 
        has not been fully captured or remains absent from these simulators. 
        The inability to simulate I/O with page cache leads to inaccurate 
        simulations in many situations, especially when dealing with data-intensive applications.

        In this paper, we propose a I/O simulation model that characterizes
        the key features of the page cache. We have implemented the model
        with SimGrid, a simulation framework for distributed systems, using
        WRENCH, a SimGrid-based workflow simulator. Our simulator is
        compatible with both single-threaded and multithreaded
        applications, and can simulate both writeback and writethrough
        cache that can factor into local and network-based filesystems. We
        evaluate it in different conditions, including sequential and
        concurrent applications, as well as local and remote I/Os.
        The page cache model reduces the simulation
        error by up to an order of magnitude compared to cacheless simulations.

@@

        The Linux Page Cache plays an important role in
        the reduction of filesystem data transfer costs. With the page cache, previously
        read data can be reread directly from memory and written data can be written to
        memory before being asynchronously flushed to disk, resulting in improved I/O performance
        on slower storage devices. The performance benefits that can be obtained from using the page
        cache are determined by many factors including the total amount of memory,
        the amount of data being written (i.e. dirty data), and the percentage of available memory available for
        written data. All these factors are important when determining the impact of data transfers on
        application performance, particularly in data intensive applications.

        Characterizing the performance of applications directly on computing infrastructure
        has various shortcomings, which include, limited and shared resources, operational
        costs, unstable environments, and the difficulties in producing reproducible results.
        Simulations address these concerns by providing abstractions for the performance of computer
        hardware, such as CPU, network and storage. Consequently, simulations provide a
        cost-effective, fast, easy and reproducible  way evaluate application performance
        on different infrastructures without having to execute the code.

        The amount of data-intensive applications has been on the rise as a result of
        open-data and data sharing initiatives. Due to the sheer size of the data being
        processed, these applications must be executed on large scale infrastructure
        such as High Performance Computing (HPC) clusters or the cloud. Determining
        which type of infrastructure is the most cost-effective for the particular
        application can be tricky without the help of a simulator.

        Page cache simulation is crucial when simulating data intensive applications,
        which rely on cache hits and writing to memory for additional performance.
        Furthermore, simulating the page cache enables us to study important application-I/O
        bottlenecks which can arise over the course of execution, such as page cache
        starvation~\cite{zhuang2017}.
        While existing simulators can adequately characterize many different computing environments,
        they lack much of the ability to adequately describe the page cache. Some
        simulators try to simulate I/O with page cache, but fail to capture the key features such
        as dirty data and cache eviction policies~\cite{nunez2012simcan,nunez2012icancloud}. 
        Other simulators, such as~\cite{xu2018saving}, capture the important features of the
        page cache, but are domain-specific.

@@

        In this paper, we present WRENCH-cache, a simulation model of I/O with page
        cache incorporated into WRENCH~\cite{casanova2018wrench}, a workflow management system simulator based on
        the popular SimGrid framework. The goal of WRENCH-cache is to provide users
        with an accurate representation of a workflow execution, when file system page
        cache is used, while providing a good balance between accuracy and scalability.
        WRENCH-cache is able to simulate both writeback and writethrough
        cache which would normally be available to local and NFS-based filesystems.
        WRENCH-cache can provide accurate simulations for large-scale parallel
        applications using the page cache.

>>>>

%\title{Simulating the Linux page cache in distributed applications with WRENCH}
\title{Simulating the Linux page cache in distributed applications with WRENCH}

\author{\IEEEauthorblockN{Hoang-Dung Do$^1$, Val\'erie Hayot-Sasson$^1$, Rafael Ferreira da Silva$^3$, Henri Casanova$^2$, Tristan Glatard$^1$
  }\\
  \IEEEauthorblockA{
    $^1$Department of Computer Science and Software Engineering, Concordia University, Montreal, Canada\\
     $^2$Department of Information and Computer Sciences, University of Hawai`i at M\=anoa, USA\\
     $^3$Department of Computer Science, University of Southern California, Los Angeles, USA
  }
}

@@

        % Henri Rewrote quite  a bit in the next paragraph here to spell out a bit
        % more why we need simulation
	The emergence of Big Data in recent years has resulted in a growing
	need for efficient data storage management techniques.  While
	infrastructures with sufficient compute power are available to
	process large datasets, the I/O bottleneck remains.  A
	well-known and ubiquitous approach for reducing the overhead of
	I/O operations is to use the Linux page cache.  It is thus crucial
	to study experimentally the impact of the Linux page cache on the
	performance Big Data applications. Given the difficulties with and shortcomings of
	real-world experiments, a popular approach for application
	performance studies is to use simulation.  While many simulation
	frameworks have been developed for this purpose, they do not 
        simulate page caching fully, or even at all.  As a result, using
	these frameworks for studying the performance of data-intensive
	applications leads inaccurate simulation results.

        % Some cosmetic rewrites below
	In this paper, we propose a I/O simulation model that characterizes
	the key features of the page cache. We have implemented this model
	as  part of the WRENCH workflow simulation framework, which itself
	builds on the popular SimGrid distributed systems simulation
	framework.  Our model and its implementation enable the simulation
	of both single-threaded and multithreaded applications, and of both
	writeback and writethrough caches for local or network-based
	filesystems. We evaluate the accuracy of our model in different
	conditions, including sequential and concurrent applications, as
	well as local and remote I/Os.  We find that our page cache model
	reduces the simulation error by up to an order of magnitude when
	compared to state-of-the-art, cacheless simulations.

@@

        The Linux Page Cache plays an important role in
        reducing filesystem data transfer times. With the page cache, previously
        read data can be re-read directly from memory and written data can be written to
        memory before being asynchronously flushed to disk, resulting in improved I/O performance
        on slower storage devices. The magnitude of the performance benefits that can be obtained from using the page
        cache are determined by many factors including the total amount of memory,
        the amount of data being written (i.e., dirty data), and the percentage of available memory available for
        written data. All these factors are important when determining the impact of I/O on
        application performance, particularly in data-intensive applications.

        % Swapped the order the two paragraphs and merged them into one
        The number of data-intensive applications has been steadily rising as a result of
        open-data and data sharing initiatives. Due to the sheer size of the data being
        processed, these applications must be executed on large scale infrastructure
	such as High Performance Computing (HPC) clusters or the cloud.  It
	is thus crucial to quantify the performance of these applications
	on these platforms. The goals include determining which type of hardware/software
	stacks are best suited to different application classes, as well as
	understanding the shortcomings of current algorithms, designs and
	technologies so as to propose new solutions. Unfortunately, application performance
        studies via experiments on real-world compute platforms
        faces several difficulties (high operational costs, labor-intensive experimental setups,
        shared platforms with dynamic loads that hinder reproductibilty of results) and shortcomings
        (experiments are limited to the available platform/softare configurations, which precludes
        the exploration of hypothetical scenarios). 
	Simulations address these concerns by providing simulation models
	and abstractions for the performance of computer hardware, such as
	CPU, network and storage. Consequently, simulations provide a
	cost-effective, fast, easy and reproducible way to evaluate
	application performance on arbitrary platform configurations. It thus comes
        as no surprise that, considering here only the field of parallel and distributed
        computing, a large number of simulation frameworks have been developed and used
        to support research and development endeavors~\cite{LAUNDRYLISTOFREFS_THAT_HENRI_WILL_PUT_IN}.

        Page caching is a well-known and ubiquitous technique for mitigating the I/O bottleneck
        faced by data-intensive applications. As such, it is necessary to simulate page caching
        when performing performance studies in simulation.
        Furthermore, simulating the page cache makes it possible to study relevant application-I/O
        bottlenecks that can arise over the course of execution, such as page cache
        starvation~\cite{zhuang2017}.
        While existing simulation frameworks of parallel and distributed computing
        systems  capture many relevant features of hardware/software stacks of interest, 
        they lack the ability to simulate page caches adequately. Those frameworks
        that implement some page cache simulation
        fail to capture key features such
        as dirty data and cache eviction policies~\cite{nunez2012simcan,nunez2012icancloud}. 
        Some simulators, such as that in~\cite{xu2018saving}, do capture such features,
        but are domain-specific.

@@

	In this work, we present WRENCH-cache, a simulation model of I/O
	with page cache incorporated into the WRENCH simulation framework~\cite{casanova2018wrench},
	which is  based on the popular SimGrid distributed simulation framework~\cite{XXX}.
	WRENCH-cache is to make it possible to simulate 
        application execution accurately in the presence of page caching. Our specific 
        contributions are as follows:
    \begin{compactitem}
        \item A page cache simulation model that supports 
	both single-threaded and multithreaded applications, and of both
	writeback and writethrough caches for local or network-based
	filesystems;
        \item An implementation of this model as part of the WRENCH framework; 
        \item An evaluation and validation of the accuracy of our model, and of its implementation,
              for a range of applications, platforms, and page cache configurations. 
    \end{compactitem}
